Summer 2025 ToDo

### 2025-10-22 Relations/Paper/Well-foundedness.tex
Almost done with section 5

The above notions give rise to several distinct definitions of well-foundedness, given below. 
Recall the definition of an $R$-decreasing sequence from Section~\ref{sec:Definitions}:
this is a function $s : \nat \to A$ satisfying $(s(k+1),s(k)) \in R$ for all $k \in \nat$.

@@

Classically, the above notions of well-foundedness are all equivalent.  Constructively, only the first two are equivalent; the remaining definitions are not. Moreover, even individual classical definitions can be given constructive
interpretations in varying degrees of logical strength.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in certain cases.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
Indeed, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.
The restriction to $\lnot\lnot$-closed subsets implies (and is equivalent to) 
every subset $P$ being merely \emph{weakly} decidable (satisfying $\lnot \lnot P(x) \lor \lnot P(X)$ 
for all $x$).

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
These are formalized in the definition below, and in the corresponding file
\texttt{WFWeakDefinitions.agda}.

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.
\apcomment{Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

@@

% The logical relationships between all of the notions of well-foundedness defined 
% above are presented in Figure~\ref{fig:WF}.
%
\begin{figure}[hb]
\centering  
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

% Row 1: WFacc
\node[box] (WFacc) {$\WFacc$};
% \node[box, right=of WFacc] (NNWFacc) {$\lnot\lnot$WFacc};
\node[box, right=2.9cm of WFacc] (WFaccM) {$\WFaccm$};

\draw[arrow] (WFacc) -- (WFaccM);
% \draw[arrow] (NNWFacc) -- (WFaccM);
\draw[arrow, bend right] (WFaccM) to node[above, text=red] {$\accDNE$} (WFacc);

% Row 2: WFind
% \node[box, below=of WFacc] (WFind) {WFind};
% \node[box, right=of WFind] (NNWFind) {$\lnot\lnot$WFind};
% \node[box, right=of NNWFind] (WFindM) {WFind-};

% \draw[arrow] (WFind) -- (NNWFind);
% \draw[arrow] (NNWFind) -- (WFindM);

% \draw[arrow] (NNWFseq) -- (WFseqM);

% Row 5: WFminDNE -> WFminDNE-
\node[box, below=of WFacc] (WFminDNE) {$\WFminDNE$};
\node[box, below=of WFaccM] (WFminDNEM) {$\WFminDNEm$};

% Row 4: WFmin -> WFmin-
\node[box, below=of WFminDNE] (WFmin) {$\WFmin$};
\node[box, below = of WFminDNEM] (WFminM) {$\WFminm$};

% Row 3: WFseq
\node[box, below=of WFmin] (WFseq) {$\WFseq$};
% \node[box, right=of WFseq] (NNWFseq) {$\lnot\lnot$WFseq};
\node[box, below= of WFminM] (WFseqM) {$\WFseqm$};

% Row 6: WFcor 
\node[box, below=of WFseq] (WFCor) {$\WFcor$};
\node[box, below=of WFseqM] (WFminCor) {$\WFcorm$};

% Arrows between rows
% \draw[arrow, <->] (WFacc) -- (WFind);
% \draw[arrow, <->] (NNWFacc) -- (NNWFind);
% \draw[arrow, <->] (WFaccM) -- (WFindM);

\draw[arrow, bend right = 90] (WFacc) to node[pos = 0.75,left, text = red] {$\rdec$} (WFseq);
\draw[arrow, bend right] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right = 90] (WFminCor) to node[right, text=red] {$\accCor$} (WFaccM);  
% \draw[arrow, bend left = 55] (WFaccM) to (WFminM);

\draw[arrow] (WFmin) -- (WFseq);
\draw[arrow] (WFminM) -- (WFseqM);
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend right] (WFminM) to (WFminDNEM);
\draw[arrow, bend right] (WFminDNEM) to (WFminM);

\draw[arrow] (WFminDNE) to node[left, text = red] {$\accDNE$} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.5, left, text=red] {$\mpe$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[right, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {$\mpe$} (WFseqM);
\draw[arrow] (WFCor) -- (WFminCor);
\draw[arrow, bend left = 110] (WFCor) to node[left, text=red] {$\accCor$} (WFacc);
\draw[arrow, bend left] (WFminCor) to node[below, text=red] {$\corDNE$} (WFCor);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
 \label{fig:WF} 
\end{figure}
\clearpage

\begin{figure}[hb]
\centering  
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

\node[box, below=of WFminDNE] (WFacc) {$\WFacc$};

\node[box, below = of WFminDNEM] (WFaccM) {$\WFaccm$};
\draw[arrow, bend right] (WFacc) to (WFaccM);
\draw[arrow, bend right] (WFaccM) to node[below, text=red] {$\accDNE$} (WFacc);

@@

% Inter-row arrows
\draw[arrow] (WFacc) to node[right, text = red] {$\rdec$} (WFseq);
\draw[arrow, bend right] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right = 90] (WFminCor) to node[right, text=red] {$\accCor$} (WFaccM);

\draw[arrow, bend right = 90] (WFmin) to (WFseq);
\draw[arrow] (WFaccM) -- (WFseqM);
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend right] (WFminM) to (WFminDNEM);
\draw[arrow, bend right] (WFminDNEM) to (WFminM);

\draw[arrow] (WFminDNE) to node[right, text = red] {$\accDNE$} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.5, left, text=red] {$\mpe$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[left, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {$\mpe$} (WFseqM);
\draw[arrow] (WFCor) -- (WFminCor);
% \draw[arrow, bend left = 110] (WFCor) to node[left, text=red] {$\accCor$} (WFacc);
\draw[arrow, bend left] (WFminCor) to node[below, text=red] {$\corDNE$} (WFCor);

\draw[arrow, dashed, bend left=80, blue] (WFCor) to node[left, text=blue] {$\accCor$} (WFacc);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
 \label{fig:WF2} 
\end{figure}
\clearpage

All of the above definitions of well-foundedness are classically equivalent.
But constructively, the only provable equivalence is between the inductive and accessible 
formluations (reflecting the fact that accessibility is defined as the least inductive predicate).
Formalizing the remaining implications enabled us to identify where exactly classical logic 
is used, and how much of it is truly needed.

The logical relationships between the various definitions are summarized in Figure \ref{fig:WF}.
An arrow from one node to another with no label indicates direct logical implication,
provable constructively with no additional assumptions. An arrow with a label indicates an additional 
(semi-)classical hypothesis needed to establish the implication. 
The definitions of these hypotheses are given in Figure \ref{tab:cprop}.
All of the implications have been 
formalized within the WellFounded subdirectory of the Agda repository.

@@

\apcomment{Make another version of the diagram, swapping WFmin and WFacc rows.}

% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
Figure \ref{tab:cprop} illustrates that the network of logical relationships between
constructive notions of well-foundedness is quite rich indeed. 
Taking the $\lnot\lnot$-closures of 
well-foundedness properties cuts down some, but not all, of the complexity.
For example, the two minimality principles become equivalent.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply 
the (weak) accessible one without an additional hypothesis $\accCor$, 
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
element that is \emph{not} strongly normalizing to a one-step reduct of it having the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

Other observations that can be gleaned from the figure include the following.
\apcomment{The macros have to fixed throughout.}
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable. 
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed 
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent, are directly implied by $\isWFaccm$, 
    and directly imply $\isWFseqm$  \apcomment{Fix macros}.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\isWFaccm$ assuming the 
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\isWFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.
    \apcomment{Add the ``is''!}

    (This is analogous to how $\isWFminDNE$ is a natural weakening of 
    $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)

@@

In \cite{Terese}, well-foundedness is defined as in the classical mathematics textbooks,
as the absence of an infinite $R$-decreasing sequence.
(Again, well-foundedness of $R$ corresponds to strong normalization of its converse.)
\apcomment{Add a discussion about isWFseq- vs. isWFseq$\lnot\lnot$.}

We consider $\isWFacc$ the canonical constructive definition of a well-founded relation.

>>>>

The above notions give rise to several distinct definitions of well-foundedness, given below.
Recall the definition of an $R$-decreasing sequence from Section~\ref{sec:Definitions}:
this is a function $s : \nat \to A$ satisfying $R(s(k+1))(sk) \in R$ for all $k \in \nat$.

@@

Classically, the above notions of well-foundedness are all equivalent.  Constructively, only the first two are equivalent; the remaining definitions are not. Moreover, even individual classical definitions can be given constructive
interpretations in varying degrees of logical strength.
Formalizing the remaining implications enabled us to identify where exactly classical logic
is used, and how much of it is truly needed.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in certain cases.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
The file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis $\WFmin$ for the standard strict order on the natural numbers entails decidability of every subset of $\nat$ (which is clearly incompatible with effective semantics), while its restriction $\WFminDNE$ implies
%(and is equivalent to)
that each subset $P$ it \emph{weakly} decidable,
satisfying $\lnot \lnot P(x) \lor \lnot P(X)$ for all $x$.

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
These weaker notions are formalized in the definition below, and in the corresponding file
\texttt{WFWeakDefinitions.agda}.

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.  This definition appears to be strictly weaker than
WFseqm-not-not, asserting that every sequence not-not contains a
non-decreasing index.
\apcomment{fix the macro please!}
% \apcomment{Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

@@

The logical relationships between all of the above definitions are summarized in Figure \ref{fig:WF}.
All of the implications in the figure have been
formalized within the WellFounded subdirectory of the Agda repository.
An arrow from one node to another with no label indicates direct logical implication,
provable constructively with no additional assumptions. An arrow with a label indicates an additional
(semi-)classical hypothesis needed to establish the implication.
The definitions of these hypotheses are given in Figure \ref{tab:cprop}.

We consider $\WFacc$ to be the canonical constructive definition of
a well-founded relation (hence the emphasis in the figure).
In the ARS setting, this corresponds to defining
strong normalization inductively in terms of reductions to normal forms.
As can be seen from the figure, it is implied by
Terese's definition $\WFseqm$ only when \emph{two}
strongly classical conditions are in place.
These are $\accDNE$ (not-not-closure of strong normalization)
and $\accCor$ (which is related to the existence of an effective perpetual reduction strategy).

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

\node[box, line width = 0.5mm, below=of WFminDNE] (WFacc) {$\WFacc$};

\node[box, below = of WFminDNEM] (WFaccM) {$\WFaccm$};
\draw[arrow] (WFacc) to (WFaccM);
\draw[arrow, bend left] (WFaccM) to node[below, text=red] {$\accDNE$} (WFacc);

@@

% Inter-row arrows
\draw[arrow] (WFacc) to node[right, text = red] {$\rdec$} (WFseq);
\draw[arrow, bend right] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right = 90] (WFminCor) to node[pos = 0.4, right, text=red] {$\accCor$} (WFaccM);

\draw[arrow, bend right = 75] (WFmin) to (WFseq);
\draw[arrow, bend left = 75] (WFminM) to (WFseqM);
\draw[arrow] (WFaccM) -- (WFseqM);
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend right] (WFminM) to (WFminDNEM);
\draw[arrow, bend right] (WFminDNEM) to (WFminM);

\draw[arrow] (WFminDNE) to node[right, text = red] {$\accDNE$} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.3, left, text=red] {$\mpe$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[left, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {$\mpe$} (WFseqM);
\draw[arrow] (WFCor) -- (WFminCor);
% \draw[arrow, bend left = 110] (WFCor) to node[left, text=red] {$\accCor$} (WFacc);
\draw[arrow, bend left] (WFminCor) to node[below, text=red] {$\corDNE$} (WFCor);

\draw[arrow, bend left=80] (WFCor) to node[pos = 0.4, left, text=red] {$\accCor$} (WFacc);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
 \label{fig:WF}
\end{figure}
% \clearpage

% All of the above definitions of well-foundedness are classically equivalent.
% But constructively, the only provable equivalence is between the inductive and accessible
% formluations, reflecting the fact that accessibility is defined as the least inductive predicate.
% Formalizing the remaining implications enabled us to identify where exactly classical logic
% is used, and how much of it is truly needed.

@@

% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.
Figure \ref{fig:WF} illustrates that the network of logical relationships between
constructive notions of well-foundedness is quite rich indeed.
Taking the $\lnot\lnot$-closures of
these notions cuts down some, but not all, of the complexity.
For example, the two minimality principles become equivalent.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply
the (weak) accessible one without an additional hypothesis $\accCor$,
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
element that is \emph{not} strongly normalizing to a one-step reduct of it having the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

Other observations that can be gleaned from the figure include the following.
\apcomment{The macros have to fixed throughout.}
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable.
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent, are directly implied by $\isWFaccm$,
    and directly imply $\isWFseqm$  \apcomment{Fix macros}.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\isWFaccm$ assuming the
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\isWFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.
    \apcomment{Remove the ``is''!}

    \apcomment{Remove?} (This is analogous to how $\isWFminDNE$ is a natural weakening of
    $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)

@@
